---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
 Knit to HTML so that you can see the clean version 


```{r}
library(dplyr)
library(gapminder)
data ("gapminder")
attach(gapminder)

head(gapminder)

```
Loaded the gapminder data, now manipulate in dplyr

```{r}
gap<-gapminder
rm(gapminder)

mean(gap[gap$continent=="Africa", "gdpPercap"]$gdpPercap) #averaging the gdpPercap columns with Africa
mean(gap$gdpPercap[continent=="Africa"]) #same but shorter code

```
```{r}
#####work with data in dplyr

ncol(gap)
names(gap)

subset_gap <- gap %>% select(year, country, gdpPercap) #only selects those columns, all rows
subset_gap

```

select() subsets data columns


filter() subsets data rows

```{r}
year_country_gdp_euro <-gap %>%
  filter(continent=="Europe") %>%
  select(year, country, gdpPercap)

year_country_gdp_euro

```

Write a command with pipes that filters the gapminder dataset to only include data from 2007 in Africa, and then select the year, country, and lifeExp columns. 
```{r}
names(gap)
practice <-gap %>%
  filter(year==2007, continent == "Africa") %>% #selects all rows where Year is 2007 and continent is Africa
  select(year, country, lifeExp) #selects these columns

dim(practice)

```


group_by() subsets your data by chunking it based on a column
turns a single data frame into a set of data frames subset by some variable


```{r}
head(gap)
print(names(gap))

str(gap)

gap_by_continent <- gap %>% group_by(continent) #group by continent
str(gap_by_continent)

#returns the mean gdp per capita for each continent
gdp_by_continent <- gap_by_continent %>%
  summarize(mean_gdpPercap=mean(gdpPercap))





```

Let's compute the average life expectancy across all African countries by year.  In how many years did average African life expectancy decrease?

```{r}

{average_life_exp <-gap %>%
  filter(continent == "Africa") %>% #selects all rows where continent is Africa
  group_by(year) %>%
  summarize(mean_lifeExp=mean(lifeExp)) 
  print(average_life_exp)}


```

Average life exp per country, which has longest and shortest?

```{r}

life_exp_avg<-gap%>%
  group_by(country)%>%
  summarize(life_exp=mean(lifeExp))

life_exp_avg%>% filter(life_exp ==min(life_exp)) #returns Sierra Leone, lowest life exp

life_exp_avg%>% filter(life_exp==max(life_exp)) #returns Iceland

```

mutate() allows you to add a column to a data frame
Note: you can also do
datafram$newcolumn<-newcolumn

```{r}
bill_gdp<- gap%>% 
  filter(year==2007)%>%
  mutate(billion_gdp = gdpPercap * pop /10^9) %>%
  select (continent, country, billion_gdp)
print(bill_gdp)

```




```{r}
std_er<-function(sample_size, sd){
  se <- (sd/sqrt(sample_size))
  return(se)
}

gdp_by_cont<- gap %>%
  group_by(continent, year) %>%
  summarize(mean_gdp=mean(gdpPercap),
            sd_gdp =sd(gdpPercap),
            mean_pop=mean(pop),
            sample_size = n() )#function n counts the sample size of variable grouped by
           
sd_gdp <- sd(gdpPercap)
sample_size <- gdp_by_cont$sample_size
serr_gdp<-(sd_gdp/sqrt(sample_size))

```


Combining dplyr with ggplot2

```{r}
library(ggplot2)
gap %>% filter(continent=="Asia")%>%
  ggplot(aes(x=gdpPercap, y= lifeExp))+
  geom_point()+
  labs(x="gdp per capita", y="Life Expectancy (years)")

```



Using tidyr
on wide vs long data frames. First we open gapminder in wide format

```{r}
View(gapwide)

gap_wide<- read.csv("data/gapminder_wide.csv", stringsAsFactors = FALSE)

colnames(gap_wide)
dim(gap_wide)

```



Want to change a wide format to a long format to make R happy

```{r}
gap_long<-gap_wide %>%
  gather(obstype_year, obs_values, -continent, -country)

gap_long<- gap_long %>%
  separate(obstype_year, into = c("obs_type", "year"), 
           sep="_") #separates by the underscore symbol

gap_long$year<-as.integer(gap_long$year) #change year from a character to an integer



```
Using the gap_long dataset, summarize the mean life expectancy by continent.

```{r}
sum_life<-gap_long %>%
  group_by(continent) %>%
  filter(obs_type=="lifeExp") %>%
  summarize(mean_life = mean(obs_values))

#can also filter first and then group the data, it works either way

```



Now we will use spread to move the data apart so it makes more sense to look at
```{r}
str(gap_long)
gap_normal<- gap_long %>%
  spread(obs_type, obs_values)
str(gap_normal)

names(gap_long)



gap_normal <- gap_normal[,names(gap)]

names(gap_normal)

all.equal(gap, gap_normal) #asks R, are these two data frames equal?, NO they are not

head(gap)
head(gap_normal)


gap_normal <- gap_normal %>%
  arrange(country, continent, year)


```

for help file you can go to Help tab at top of the R studio, and click cheat sheets, dplyr and it visually shows how to use the functions







